---
title: "Functional Data Analysis Tutorial"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-background: true
    code-tools: true
    code-fold: false
    code-block-border-left: true
    theme: flatly
    css: styles.css
execute:
  echo: true
  cache: true
  message: false
  warning: false
editor: source
---

```{r load packages} 
#| include: false 
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(refund)
library(viridis)
library(mgcv)
```

This tutorial provides an introduction to several key functional data analysis methods. Note that we will use the `refund` package, `vbvs.concurrent`, and `mgcv` to do functional data analysis. We begin with functional principal components analysis, then move to scalar-on-function regression, and function-on-scalar regression.


## Pupillometer data

We will use examples based on curves of pupil response to a light stimulus. Some subjects smoke cannabis 60 minutes before the light stimulus and others do not. The goal is to assess whether the pupil response to light differs for those who used cannabis. 

You can download the data [here](Downloads/pupil.Rdata). 

```{r}
load("/Users/JWROBE8/icloud/Documents/websites/tutorials/Downloads/pupil.Rdata")

head(pupil)
```

The functional data are percent change in pupil diameter over the course of 5 seconds immediately after the pupil is exposed to a flash of light. This is referred to as the **pupil light response curve**.


Variables included in this dataset are:

- `id`: subject id
- `use`: whether a subject used cannabis, `1=use` and `0=no use`
- `age`, in years
- `alcohol`: average weekly number of alcoholic beverages
- `seconds`: the functional domain
- `percent_change`: the pupil light response curve collected 40 minutes **after** smoking cannabis for those with `use = 1` and after an equivalent no-smoking rest time for those with `use = 0`
- `percent_change_baseline`: the pupil light response curve values at baseline, before any participants have consumed cannabis



Curves are shown below:

```{r}

baseline = pupil %>%
  mutate(use = factor(use, levels = 1:0, labels = c("Cannabis use", "No use"))) %>%
  ggplot(aes(seconds, percent_change_baseline)) +
  geom_line(aes(group = id), alpha = 0.1) +
  geom_smooth(aes(color = factor(use), linetype = factor(use)), se = FALSE) +
  #facet_wrap(~use_group) +
  theme_minimal() +
  labs(x = "Seconds since light stimulus", y = "Percent change in pupil diameter") +
  scale_color_viridis(discrete = TRUE) +
  ylim(-60, 20) +
  theme(legend.position = "bottom") 

post = pupil %>%
  mutate(use = factor(use, levels = 1:0, labels = c("Cannabis use", "No use"))) %>%
  ggplot(aes(seconds, percent_change)) +
  geom_line(aes(group = id), alpha = 0.1) +
  geom_smooth(aes(color = factor(use), linetype = factor(use)), se = FALSE) +
  labs(x = "Seconds since light stimulus", y = "Percent change in pupil diameter") +
  scale_color_viridis(discrete = TRUE) +
  #facet_wrap(~use_group) +
  theme_minimal() +
  ylim(-60, 20) +
  theme(legend.position = "bottom") 
  
baseline + post
```


## FPCA

## SoFR

For scalar-on-function regression with a Gaussian outcome the model we are interested in is 

$$Y_i = \beta_0 + \sum_j \beta_jX_{ij} + \sum_k\int_t \mathcal{B}_k(t)Z_{ik}(t)dt + \epsilon_i,$$
where:

- $Y_i$ is a scalar outcome
- Each $X_{ij}$ is a scalar covariate
- Each $\beta_j$ is a regression coefficient for a scalar covariate
- Each $Z_{ik}(t)$ is a functional covariate
- Each $\mathcal{B}_k(t)$ is a *coefficient function*, a regression coefficient for a functional covariate
- $\epsilon_i$ are normally distributed iid errors

The generalized linear model (GLM) version of this model (for non-Gaussian outcomes, usually binary) is:

$$g(E[Y_i|X_i, Z_i]) = \beta_0 + \sum_j \beta_jX_{ij} + \sum_k\int_t \mathcal{B}_k(t)Z_{ik}(t)dt,$$
where $g(\cdot)$ is a known link function (for example, the logit link in logistic regression).


### Important practical considerations

* For SoFR, can't have missing functional observations so need to impute or remove
  * The default in many implementations is to use `complete.cases()` (check your sample size in the results).
* For `pfr()`, binomial responses should be specified as a numeric vector rather than as a matrix or a factor.


### refund::pfr()

First, we fit a model with a continuous scalar outcome (age) and a functional predictor (percent change in pupil size) using the `pfr()` function from the `refund` package. pfr stands for "penalized functional regression".


The data need to be set up in a specific way for modeling, so we will process the data first.

```{r}
sofr_df = pupil %>%
  select(-percent_change_baseline) %>%
  pivot_wider(names_from = seconds, values_from = percent_change, 
                names_prefix = "t_") %>%
  as.data.frame()

pupil_mat = sofr_df %>% select(starts_with("t_")) %>% as.matrix()

# reorganize data for use with refund::pfr() function
sofr_df$percent_change = pupil_mat
```


```{r}
pfr_age = pfr(age ~ lf(percent_change, k = 30, bs = "cr"),
              data = sofr_df)
```


```{r}
summary(pfr_age)

plot(pfr_age)
```




```{r}
pfr_use = pfr(use ~ lf(percent_change, k = 30, bs = "cr") + age + alcohol,
              family = binomial,
              data = sofr_df)
```

Below we show the model summary and default plotting function.

```{r}
summary(pfr_use)
plot(pfr_use)
```



```{r, echo = FALSE, eval = FALSE}
# You can also extract and plot the coefficient function and its 95% C.I. yourself:
coef(pfr_use) %>%
  rename(time = percent_change.argvals,
         beta_hat = value) %>%
  mutate(lower = beta_hat - 1.96 * se,
         upper = beta_hat + 1.96 * se) %>%
    ggplot(aes(time, beta_hat)) + 
  geom_line() +
  geom_line(aes(y = lower), linetype = 2, color = "blue") +
  geom_line(aes(y = upper), linetype = 2, color = "blue") +
  geom_hline(yintercept = 0, linetype = 3, color = "red") +
  theme_minimal()
```


### mgcv::gam()

You can also fit this model directly using `mgcv::gam()` First, we need to process the data a little bit differently—some of this is done under the hood in `pfr()`.

```{r}
ncols = ncol(pupil_mat)

# vector, matrix containing functional domain of observed data (assumes everyone is on the same grid)
sind = seq(0, 1, len = ncols)
smat = matrix(sind, nrow(sofr_df), ncols, byrow = TRUE)
  
# construct quadrature weights for numeric integration  
sofr_df$smat = I(smat)
sofr_df$lmat = I(matrix(1/ncols, nrow(sofr_df), ncols))
sofr_df$zlmat = I(sofr_df$lmat * sofr_df$percent_change)


```


```{r, eval = FALSE, echo = FALSE}
# see FDA book page 131 for constructing quadrature weights when domain is not on 0, 1
#s_vec = unique(pupil$seconds)
#smat = matrix(s_vec, nrow(sofr_df), ncols, byrow = TRUE)

# calculate vector quadrature weights using Simpson's rule 

```


```{r}
gam_use = gam(use ~ s(smat, by=zlmat, bs = "cr", k = 30) + age + alcohol, 
                   data= sofr_df,
                   method = "REML", family = binomial)
```



```{r}
summary(gam_use)
plot(gam_use)
```



#### Plotting in mgcv


Here we show how to obtain pointwise confidence intervals for coefficient functions


```{r}

# put required data inputs into a dataframe
# need to have same names as covariates in model
s_pred = seq(0,1, length.out = 100)
df_pred = data.frame(smat = s_pred, zlmat = 1, alcohol = 0, age = 0)

# call predict.gam
coef_est = predict(gam_use, newdata = df_pred, type = "terms", se.fit = TRUE)

# extract point estimate and pointwise standard errors
beta_hat = coef_est$fit[,3]
se_beta_hat = coef_est$se.fit[,3]

# construct lower and upper 95% confidence bounds
lower = beta_hat - 1.96 * se_beta_hat
upper = beta_hat + 1.96 * se_beta_hat
```


We now plot the results. We are plotting $\beta(t)$ here, but we could also plot $e^{\beta(t)}$ to obtain the odds ratio.

```{r}
tibble(time = s_pred, beta_hat = beta_hat, lower = lower, upper = upper) %>%
  ggplot(aes(time, beta_hat)) + 
  geom_line() +
  geom_line(aes(y = lower), linetype = 2, color = "blue") +
  geom_line(aes(y = upper), linetype = 2, color = "blue") +
  geom_hline(yintercept = 0, linetype = 3, color = "red") +
  theme_minimal()
```


## FoSR

The function-on-scalar regression model is 

$$Y_i(t) = \beta_0(t) + \sum_j \beta_j(t)X_{ij} + b_i(t) + \epsilon_i(t),$$

where:

- $Y_i(t)$ is a functional outcome
- Each $X_{ij}$ is a scalar covariate
- Each $\beta_j(t)$ is a coefficient function
- $b_i(t)$ is a subject-specific functional random effect. This captures correlation within subjects over time that is not captured by the mean. This term is not always included in FoSR models, but it is generally a good idea because it yields better inference.
- $\epsilon_i(t)$ are independent, normally distributed errors.


### Important practical considerations

* You need to dummy-code categorical variables (e.g., 1 vs. 0 for binary variables).
* The subject ID variable should be a factor.
* Pointwise confidence intervals are technically subject to multiple comparisons issues, and joint bands or bootstrap-based approaches can be preferred.




### Modeling

For FoSR, we can use `mgcv::gam()` directly. This modeling process happens in 3 steps:

1. Fit the mean model $Y_i(t) = \beta_0(t) + \sum_j \beta_j(t)X_{ij} + \epsilon_i(t)$, assuming iid errors across individuals and the domain.
2. Obtain residuals from the model in step (1) and decompose these into principle directions of variation (eigenfunctions) using FPCA.
3. Refit the model using top eigenfunctions from step (2) as random effects. This step allows us to estimate $b_i(t)$ and obtain more valid inference for our coefficient function.


These steps are implemented below:

```{r}
##########################################################################################
## step 1
mean_mod = mgcv::gam(percent_change ~ s(seconds, k=30, bs="cr") + 
                          s(seconds, by=use, k=30, bs = "cr"), 
                  data = pupil, method = "REML")

########################################################################################## 
## step 2
## Create a matrix of residuals

resid_df = pupil %>%
  filter(!is.na(percent_change)) %>%
  select(id, seconds) %>%
  mutate(resid = mean_mod$residuals) %>%
  pivot_wider(names_from = seconds, values_from = resid, names_prefix = "resid.")
  
resid_mat = as.matrix(resid_df[,-1])
rownames(resid_mat) = resid_df$id

fpca_results = fpca.face(resid_mat, argvals = unique(pupil$seconds), knots = 15)
eigenfunctions <- as.data.frame(fpca_results$efunctions)
colnames(eigenfunctions) <- paste0("Phi", seq(1, fpca_results$npc))
eigenfunctions$seconds <- unique(pupil$seconds)
pupil = pupil %>% left_join(., eigenfunctions, by = "seconds") %>%
  as_tibble() %>%
  arrange(id, seconds) %>%
  mutate(id = factor(id))

##########################################################################################
# Step 3

fosr_mod <- mgcv::bam(percent_change ~ 
                            s(seconds, k=30, bs="cr") + 
                            s(seconds, by=use, k=30, bs = "cr") + 
                            s(id, by = Phi1, bs="re") + 
                            s(id, by = Phi2, bs="re")+
                            s(id, by = Phi3, bs="re") + 
                            s(id, by = Phi4, bs="re"), 
                          method = "fREML", data = pupil, discrete = TRUE)
summary(fosr_mod)

```


### Plotting

Here we will plot results much the same way we did when plotting results for SoFR.

```{r}
# put required data inputs into a dataframe
# need to have same names as covariates in model
s_pred = seq(0,1, length.out = 100)
df_pred = pupil %>%
  filter(id == first(id))

# call predict.gam
coef_est = predict(fosr_mod, newdata = df_pred, type = "terms", se.fit = TRUE)



tibble(beta1 =  coef_est$fit[, 2],
       seconds = df_pred$seconds,
       se_beta1 = coef_est$se.fit[,2],
       lower = beta1 - 1.96 * se_beta1,
       upper = beta1 + 1.96 * se_beta1) %>%
    ggplot(aes(seconds, beta1)) + 
  geom_line() +
  geom_line(aes(y = lower), linetype = 2, color = "blue") +
  geom_line(aes(y = upper), linetype = 2, color = "blue") +
  geom_hline(yintercept = 0, linetype = 3, color = "red") +
  theme_minimal()


```







## Resources

- I highly recommend this [FDA with R book](https://www.amazon.com/Functional-Analysis-Monographs-Statistics-Probability/dp/1032244712/ref=sr_1_6?crid=UEKJYVS8EV6R&keywords=crainiceanu&qid=1698108861&sprefix=crainiceanu%2Caps%2C83&sr=8-6)


Some papers about models covered in this tutorial:

- [Functional data analysis with pupillometer data](https://karger.com/dib/article/8/1/83/906653)
- [SoFR](https://pmc.ncbi.nlm.nih.gov/articles/PMC3285536/)
- [Functional linear concurrent model](https://pmc.ncbi.nlm.nih.gov/articles/PMC5457356/)
- [FoFR](https://pmc.ncbi.nlm.nih.gov/articles/PMC4560367/pdf/nihms-589200.pdf)




